---
title: "Stats Challenge"
author: "Simon and Doris"
date: "May 30, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
rm(list=ls())
library(dplyr)
library(ggplot2)
library(gdata)

```

# Data and research question

Summary:
The authors studied the influence of weaning age in cats regarding different characteristics, such as shyness towards people or objects, aggressiveness, or stereotypical behavior.
For this, they published questionnaires online for owners to answer.
It is a descriptive research design (survey), with no controls or treatment.
They grouped cats together by breed or breed type and analyzed the data, correcting for environmental influences such as age, gender, whether the cats were neutered, and whether they went outdoors.

Our approach:
reproduce analysis of Table 1 with the test used in the publication, redo with different statistical test

Original article: https://www.nature.com/articles/s41598-017-11173-5
Data: https://figshare.com/articles/Salonen_et_al_Breed_differences_of_heritable_behaviour_traits_in_cats_-_data/8143835


```{r }
data <- read.csv(file="publisheddata-scientificreports.csv", header=TRUE, sep=";")

```

Get an overview of the data

```{r}
count(data)

histrv<-hist(data$Age, main="Age distribution of cats",  breaks=25)
hist_cats <- histrv$counts #give numbers of how many per break (year)

median(data$Age)
mean(data$Age)

hist(data$Gender)
breed_count <- count(data, Breed_group)
ggplot(breed_count, aes(x=Breed_group, y=n))+
geom_bar(stat = "identity", color="grey")

weaning_count <- count(data, Weaning_age)
ggplot(weaning_count, aes(x=Weaning_age, y=n))+
geom_bar(stat = "identity", color="grey")

aggr_count <- count(data, Aggression_owner)
ggplot(aggr_count, aes(x=Aggression_owner, y=n))+
geom_bar(stat = "identity", color="grey")

```

Describe presence or absence of aggression as binary value

```{r}
data$aggression_bin <- rep(0, count(data))

data$aggression_bin[data$Aggression_owner >1]=1
aggr_count_bin <- count(data, aggression_bin)

#aggression towards owner, all
p1 <- ggplot(aggr_count, aes(x=Aggression_owner, y = n))+
 geom_bar(stat = "identity", fill = c("#32CD32", "red", "red", "red", "red"))+
 scale_x_continuous(breaks = c(1,2,3,4,5),
                    labels = c("no aggression","a little","moderately","a lot",
                               "very much") )+
 ggtitle(" ")+
 xlab("Aggression level")+
 ylab("Number of cats")+
 theme_bw()

#aggression towards owner, bin
p2 <- ggplot(aggr_count_bin, aes(x=aggression_bin, y = n))+
 geom_bar(stat = "identity", fill = c("#32CD32", "red"))+
 scale_x_continuous(breaks = c(0,1),
                    labels = c("no aggression","aggression") )+
 ggtitle(" ")+
 xlab("Aggression level")+
 ylab("Number of cats")+

 theme_bw()

library("gridExtra")

grid.arrange(p1, p2, ncol = 2, top = "Aggression towards family members")


```

Investigate different parameters vs. aggression

```{r}
#aggression vs age
hist_cats #individuals per age group

#Weaning age
table_WA_OA=table(data$Weaning_age, data$aggression_bin)
table_WA_OA
#plot(table_WA_OA, main="Aggression towards owner")
test_w <- chisq.test(table_WA_OA)
test_row1=c(test_w$statistic[1], test_w$p.value, test_w$parameter[1])
test_row1

#gender
table_G_OA=table(data$Gender, data$aggression_bin)
table_G_OA
#plot(table_G_OA, main="Aggression towards owner")
test_G <- chisq.test(table_G_OA)
test_row2=c(test_G$statistic[1], test_G$p.value, test_G$parameter[1])
test_row2

#age --> should not be done with chi square --> ANOVA
table_A_OA=table(data$Age, data$aggression_bin)
table_A_OA
#plot(table_A_OA, main="Aggression towards owner")
test_A <- chisq.test(table_A_OA)
test_row3=c(test_A$statistic[1], test_A$p.value, test_A$parameter[1])
test_row3

data$age_bin <- rep(0, count(data))
data$age_bin[data$Age >11.983]=1

table_Abin_OA=table(data$age_bin, data$aggression_bin)
table_Abin_OA
#plot(table_Abin_OA, main="Aggression towards owner")
test_Abin <- chisq.test(table_Abin_OA)
test_row3a=c(test_Abin$statistic[1], test_Abin$p.value, test_Abin$parameter[1])
test_row3a

#Hormonal status
table_N_OA=table(data$Neuter_status, data$aggression_bin)
table_N_OA
#plot(table_N_OA, main="Aggression towards owner")
test_N <- chisq.test(table_N_OA)
test_row4=c(test_N$statistic[1], test_N$p.value, test_N$parameter[1])
test_row4

#Breed
table_B_OA=table(data$Breed_group, data$aggression_bin)
table_B_OA
#plot(table_B_OA, main="Aggression towards owner")
test_B <- chisq.test(table_B_OA)
test_row5=c(test_B$statistic[1], test_B$p.value, test_B$parameter[1])
test_row5

#Other cats
table_cat_OA=table(data$Other_cats, data$aggression_bin)
table_cat_OA
#plot(table_cat_OA, main="Aggression towards owner")
test_cat <- chisq.test(table_cat_OA)
test_row6=c(test_cat$statistic[1], test_cat$p.value, test_cat$parameter[1])
test_row6


#Extraversion
table_E_OA=table(data$Activity_level, data$aggression_bin)
table_E_OA
#plot(table_E_OA, main="Aggression towards owner")
test_E <- chisq.test(table_E_OA)
test_row7=c(test_E$statistic[1], test_E$p.value, test_E$parameter[1])
test_row7

#Shyness
table_s_OA=table(data$Shyness_component, data$aggression_bin)
table_s_OA
#plot(table_s_OA, main="Aggression towards owner")
test_s <- chisq.test(table_s_OA)
test_row8=c(test_s$statistic[1], test_s$p.value, test_s$parameter[1])
test_row8

#outdoors
table_od_OA=table(data$Outdoors, data$aggression_bin)
table_od_OA
#plot(table_Od_OA, main="Aggression towards owner")
test_od <- chisq.test(table_od_OA)
test_row9=c(test_od$statistic[1], test_od$p.value, test_od$parameter[1])
test_row9

check= data.frame("Weaning_age"=test_row1, "Sex"=test_row2, "Age"=test_row3, "Hormonal_status"=test_row4, "Breed"=test_row5, "Access_to_outdoors"=test_row9, "Other_cats"=test_row6, "Shyness"=test_row8, "Extraversion"=test_row7)
check <- t(check)
check <- as.data.frame(check)

names(check)[names(check) == "V2"] <- "p_value"
check
```
#State the null-hypothesis and the alternative hypothesis
H0: weaning age and aggression towards owner are independent
H1: they are not independent

#Corrections in tests
Do Wilcoxon-test for age and shyness component

```{r}
#for age
shapiro.test(sample(data$Age, 5000, replace=FALSE))
#not normally distributed

log_age= log2(data$Age)

hist(log_age)

#still not normal after log-transformation

wilcox.test(data$Age ~ data$aggression_bin)

boxplot(data$Age ~ data$aggression_bin, ylab="Age", xlab= "Aggression", col="light blue",data = data, outcol = "red")
ggplot(data, aes(x=data$aggression_bin, y=data$Age, group=data$aggression_bin))+geom_violin()+ggtitle("Aggression and Age") +
xlab("Aggression")+ylab("Age")


#shyness component
library(ggplot2)

shapiro.test(sample(data$Shyness_component, 5000, replace=FALSE))
hist(data$Shyness_component)

wilcox.test(data$Shyness_component ~ data$aggression_bin)

boxplot(data$Shyness_component ~ data$aggression_bin, ylab="Age", xlab= "Shyness", col="light blue",data = data, outcol = "red")

ggplot(data, aes(x=data$aggression_bin, y=data$Shyness_component, group=data$aggression_bin,))+geom_violin()+ggtitle("Aggression and Shyness component") + xlab("Aggression") + ylab("Shyness component")

```

#groups by weaning age

```{r}
#table 2

group_1 <- data[(data$Weaning_age ==1|data$Weaning_age ==4), ]

table_group_1 <- table(group_1$Weaning_age, group_1$aggression_bin)
test_group_1 <- chisq.test(table_group_1)
test_group_1

group_2 <- data[(data$Weaning_age ==2|data$Weaning_age ==4), ]

table_group_2 <- table(group_2$Weaning_age, group_2$aggression_bin)
test_group_2 <- chisq.test(table_group_2)
test_group_2

group_3 <- data[(data$Weaning_age ==3|data$Weaning_age ==4), ]

table_group_3 <- table(group_3$Weaning_age, group_3$aggression_bin)
test_group_3 <- chisq.test(table_group_3)
test_group_3

group_4 <- data[(data$Weaning_age ==5|data$Weaning_age ==4), ]

table_group_4 <- table(group_4$Weaning_age, group_4$aggression_bin)
test_group_4 <- chisq.test(table_group_4)
test_group_4

group_5 <- data[(data$Weaning_age <5|data$Weaning_age ==5), ]

table_group_5 <- table(group_5$Weaning_age, group_5$aggression_bin)
test_group_5 <- chisq.test(table_group_5)
test_group_5

group_6 <- data[(data$Weaning_age > 5|data$Weaning_age <=5), ]

table_group_6 <- table(group_6$Weaning_age, group_6$aggression_bin)
test_group_6 <- chisq.test(table_group_6)
test_group_6

```

#Power analysis

< 0.1 = trivial effect
0.1 - 0.3 = small effect
0.3 - 0.5 = moderate effect
> 0.5 = large difference effect

Usage

pwr.chisq.test(w = NULL, N = NULL, df = NULL, sig.level = 0.05, power = NULL)

Arguments

w: Effect size
N: Total number of observations
df:degree of freedom (depends on the chosen test)
sig.level: Significance level (Type I error probability) -> rejection of a true null hypothesis ("false positive")
power: Power of test (1 minus Type II error probability) -> failure of rejecting a false null hypothesis (also known as a "false negative")

Large studies(>2000) may have small and often inconsequential effects but be statistically significant.
https://research.usu.edu//irb/wp-content/uploads/sites/12/2015/08/A_Researchers_Guide_to_Power_Analysis_USU.pdf 

```{r}
#power analysis
#alpha: 0.05
#w = effect size : 0.2 (be able to detect a small effect)

library("pwr")

pwr.chisq.test(w=0.2, df = 7, sig.level = 0.05, power=0.9) #power analysis for df=7 as in weaning age

```
#Regression model, basic
The authors get AIC 3099.964

```{r}
#lm(data$aggression_bin ~ data$Gender + data$Weaning_age + data$Breed_group)

regrcats <- glm(data$aggression_bin ~ data$Age +data$Gender+ data$Weaning_age #+ data$Breed_group + data$Other_cats + data$Neuter_status + data$Extraversion_component + data$Outdoors+ data$Shyness_component
                ,family=  "binomial"(link = "logit")
                )

summary(regrcats)

```
#Regression model, with breed
They get AIC 3015.063

```{r}
#lm(data$aggression_bin ~ data$Gender + data$Weaning_age + data$Breed_group)

regrcats <- glm(data$aggression_bin ~ data$Age +data$Gender + data$Weaning_age + data$Breed_group+ data$Neuter_status + data$Other_cats + data$Extraversion_component
                ,family=  "binomial"(link = "logit")
                )

summary(regrcats)
```

# How to improve methodology

- better analyze "not weaned" category: includes adult cats and very young kittens alike
- explain how they transformed age into a binary
- use appropriate test (not Chi-squared)

